A Quasi-Genuine FIFO Total Order Multicast Primitive

### 2011-07-15 textos/optpaxos/draftv3.tex
finished proofs of correctness

\blankline
\INDENT{\textit{To} \cms{} \textit{a message $m$}}
  \STATE $m.ts \leftarrow (getTime(),0)$
  %\COMMENT{current wallclock value as the timestamp of $m$}
  \STATE \rmcast{$m$}{$\{G\}$} \label{algline:rmlocal}
%  \INDENT{\textbf{for all} $p' \in m.dst \cup \{G\}$ \textbf{do}}
%    \STATE send($p'$, $m$)    
%    \COMMENT{send optimistically $m$ to all involved processes}
%  \ENDINDENT
\ENDINDENT

@@

\blankline
\INDENT{\textit{When} Decide($k'$,$msgSet$)}
%  \STATE $m.gs \leftarrow k$
  \INDENT{\textbf{while} $\exists m \in msgSet : (\forall m' \in msgSet : m \neq m' \Rightarrow m.ts < m'.ts)$ \textbf{do}}
    \STATE $msgSet \leftarrow msgSet \setminus \{m\}$
    \COMMENT{the messages are handled in ascending order of timestamp}
    \IF{$\exists m' \in decided : m'.ts \geq m.ts \wedge (\nexists m'' \in decided : m''.ts > m'.ts)$}
      \STATE $m.ts \leftarrow (m'.ts.rtc, m'.ts.seq + 1)$
    \ENDIF     
    \STATE $decided \leftarrow decided \cup \{m\}$  \label{algline:addtodecided}
    \IF{$G \in m.dst$} \label{algline:checkcons}
      \STATE $barPending \leftarrow barPending \cup \{m\}$  \label{algline:insbp2}
    \ENDIF
    \STATE \rmcast{$m$}{$m.dst \setminus \{G\}$} \label{algline:rmothers}
  \ENDINDENT
  \STATE $nextProp \leftarrow k' + 1$
  \STATE $k \leftarrow k' + 1$
\ENDINDENT

@@

\begin{proof}
In Algorithm \ref{algorithm:deliveryminimal}, whenever a process in a group $G$ \cms{}s a message, it is first \rmc{} to all processes in $G$ (l. \ref{algline:rmlocal}). As the \rmc{} is reliable, every correct process from $G$ \rmd{}s $m$. Once $m$ is \rmd{}ed by each process $p \in G$, it is inserted into the $propPending$ list of that process, to be included in the next proposal from $p$. As every process from $G$ received $m$ and enqueued it for being proposed, and since $m$ is only removed from $propPending$ when decided (l. \ref{algline:removedecided}), all processes will proposed it at some point and $m$ will eventually be decided. At this point, if $G \in m.dst$, every correct process of $G$ will insert $m$ into its $barPending$ list (l. \ref{algline:insbp2}). Besides, each correct process in $G$ \rmc{}s $m$ to all other destination groups that $m$ might be adressed to.

@@

%===                 ==============================
%===   PROPOSITION   ==============================
%===                 ==============================
\begin{props}
Once a process \cms{}s a message, then one of the correct processes that is a destination of that message eventually delivers it.
\end{props}

@@

%===                 ==============================
%===   PROPOSITION   ==============================
%===                 ==============================
\begin{props}
If a correct process which is a destination of a message $m$ delivers it, then all correct processes that are also destinations of $m$ deliver it as well.
\end{props}

@@

\subsection{Uniform Total Order}

@@

>>>>

\blankline
\INDENT{\textit{To} \cms{} \textit{a message $m$}}
  \STATE $m.ts \leftarrow (getTime(),0)$ \label{algline:gettime}
  %\COMMENT{current wallclock value as the timestamp of $m$}
  \STATE \rmcast{$m$}{$\{G\}$} \label{algline:rmlocal}
%  \INDENT{\textbf{for all} $p' \in m.dst \cup \{G\}$ \textbf{do}}
%    \STATE send($p'$, $m$)    
%    \COMMENT{send optimistically $m$ to all involved processes}
%  \ENDINDENT
\ENDINDENT

@@

\blankline
\INDENT{\textit{When} Decide($k'$,$msgSet$)} \label{algline:decide}
%  \STATE $m.gs \leftarrow k$
  \INDENT{\textbf{while} $\exists m \in msgSet : (\forall m' \in msgSet : m \neq m' \Rightarrow m.ts < m'.ts)$ \textbf{do}} \label{algline:whiledecided}
    \STATE $msgSet \leftarrow msgSet \setminus \{m\}$
    \COMMENT{the messages are handled in ascending order of timestamp}
    \IF{$\exists m' \in decided : m'.ts \geq m.ts \wedge (\nexists m'' \in decided : m''.ts > m'.ts)$} \label{algline:conditionchangets}
      \STATE $m.ts \leftarrow (m'.ts.rtc, m'.ts.seq + 1)$ \label{algline:changetsafterdecision}
    \ENDIF     
    \STATE $decided \leftarrow decided \cup \{m\}$  \label{algline:addtodecided}
    \IF{$G \in m.dst$} \label{algline:checkcons}
      \STATE $barPending \leftarrow barPending \cup \{m\}$  \label{algline:insbp2}
    \ENDIF
    \STATE \rmcast{$m$}{$m.dst \setminus \{G\}$} \label{algline:rmothers}
  \ENDINDENT
  \STATE $nextProp \leftarrow k' + 1$
  \STATE $k \leftarrow k' + 1$
\ENDINDENT

@@

\begin{proof}
In Algorithm \ref{algorithm:deliveryminimal}, whenever a process in a group $G$ \cms{}s a message, it is first \rmc{} to all processes in $G$ (l. \ref{algline:rmlocal}). From the uniform agreement property of the \rmc{} primitive, every correct process from $G$ \rmd{}s $m$. Once $m$ is \rmd{}ed by each process $p \in G$, it is inserted into the $propPending$ list of that process, to be included in the next proposal from $p$. As every process from $G$ received $m$ and enqueued it for being proposed, and since $m$ is only removed from $propPending$ when decided (l. \ref{algline:removedecided}), every processes of $G$ will propose it at some point and $m$ will eventually be decided. At this point, if $G \in m.dst$, every correct process of $G$ will insert $m$ into its $barPending$ list (l. \ref{algline:insbp2}). Besides, each correct process in $G$ \rmc{}s $m$ to all other destination groups that $m$ might be adressed to.

@@

%===                 ==============================
%===   PROPOSITION   ==============================
%===                 ==============================
\begin{props} \label{props:validity}
Once a process \cms{}s a message, then all correct processes that are destinations of that message eventually deliver it.
\end{props}

@@

%===                 ==============================
%===   PROPOSITION   ==============================
%===                 ==============================
\begin{props} \label{props:agreement}
If a correct process which is a destination of a message $m$ delivers it, then all correct processes that are also destinations of $m$ deliver it as well.
\end{props}

@@

\subsection{Atomic Order}

%===           ====================================
%===   LEMMA   ====================================
%===           ====================================
\begin{lems} \label{lemma:agreetimestamps}
If two correct processes are both destinations of a message $m$, then they agree on its final timestamp $m.ts$.
\end{lems}

\begin{proof}
Let $p$ be the process that is \cms{}ing $m$. We can prove this lemma by demonstrating that any other process $p'$ will agree with $p$ on the timestamp of $m$. Let us first consider the case of $p'$ belonging to the same group as $p$. We can prove by induction on the identifier $k$ of each consensus instance within $goup(p)$:\\

\noindent Base case ($k=1$): As this is the first consensus instance of the group, it means that the $decided = \emptyset$. From the agreement property of consensus, we know that $p$ and $p'$ decided the same contents for $msgSet$ \mbox{(l. \ref{algline:decide})}. Each message included in such agreed set also includes an initial timestamp field. As the $decided$ set is empty and, therefore (l. \ref{algline:conditionchangets}) this timestamp is not changed in neither $p$ or $p'$, both processes consider the same timestamp value for every message. Besides, as the $msgSet$ is the same for both, the $decided$ remains identical for both $p$ and $p'$.\\

\noindent Induction step: Suppose that $p$ and $p'$ have already learnt the decisions of instance $k$, their $decided$ sets remained identical and they decided the same timestamp value of each message sent from some process in their group so far. As we assume that all processes within a group learn the decided values of the different consensus instance in the same order, then both $p$ and $p'$ will next learn the decision of the same instance $k+1$. From the agreement property of consensus, the $msgSet$ decided is the same for both processes. In the Algorithm \ref{algorithm:deliveryminimal}, the timestamps may be changed after the consensus decision only in line \ref{algline:changetsafterdecision}, based on what is already in the $decided$ set (l. \ref{algline:conditionchangets}). As the $msgSet$ and $decided$ sets are each identical in $p$ and $p'$, they will make the exact same change to the timestamp of each message in the \textit{while} loop (l. \ref{algline:whiledecided} to l. \ref{algline:rmothers}). Therefore, they also agree on the timestamps of each message decided on consensus instance $k+1$ and their $decided$ sets remain identical.\\

Now, consider the case of $p'$ not belonging to the same group of $p$. Let $q$ be any process of $group(p)$. After setting the final timestamp of $m$ (l. \ref{algline:whiledecided} to l. \ref{algline:rmothers}), $q$ \rmc{}s $m$ to the $group(p')$ (l. \ref{algline:rmothers}). Once $p'$ \rmd{}s $m$, it never changes the timestamp of $m$, which already contains the final value, set by $q$. As any two processes on the same group from which $m$ has been multicast agree on its timestamp, and since any process from other group which is also a destination of $m$ agree on its timestamp as well, than any two processes that are destinations of $m$ agree on its timestamp.
\end{proof}

%===           ====================================
%===   LEMMA   ====================================
%===           ====================================
\begin{lems} \label{lemma:respecttimestamps}
If two messages $m$ and $m'$ have both a correct process $p$ as a destination, and $m.ts < m'.ts$, then $p$ delivers $m'$ after delivering $m$.
\end{lems}

\begin{proof}
Consider Algorithm \ref{algorithm:deliveryminimal}. Suppose, by means of contradiction, that $m'$ has already been delivered by $p$, whereas $m$ has not. Message $m$ might have already been inserted into $barPending$ or not. In the former case, $m'$ could not have been delivered, since $m.ts < m'.ts$, thus it would not satisfy the condition from line \ref{algline:inbarpending} until $m$ has been delivered, so we have a contradiction in the case that $m$ was already in $barPending$.

The other case is if $m$ has not been inserted into $barPending$. From line \ref{algline:inbarpending}, we know that $p$ has received a some barrier $b$ from every group in $sendersTo(group(p))$. Therefore, from Lemma \ref{lemma:oncedecidednotsmaller}, and since a barrier is also a message, we know that any new message that arrives at $p$ will have a timestamp greater than $m'.ts$. As $m$ has arrived after $m'$, then $m.ts > m'.ts$, which is also a contradiction.

We have proven that the timestamp order will not be violated by $p$. Considering that, once a message has been multicast, it will be delivered (Proposition \ref{props:validity}), then we know that all messages will be delivered by $p$, and in the correct timestamp order.\end{proof}

%===                 ==============================
%===   PROPOSITION   ==============================
%===                 ==============================
\begin{props} \label{props:atomicorder}
If processes $p$ and $p'$ are both in $m.dst$ and $m'.dst$, then $p$ delivers $m$ before $m'$ if, and only if, $p'$ delivers $m$ before $m'$.
\end{props}

\begin{proof}
Immediate from Lemma \ref{lemma:agreetimestamps}, Lemma \ref{lemma:respecttimestamps} and Proposition \ref{props:agreement}.
\end{proof}

@@

%===           ====================================
%===   LEMMA   ====================================
%===           ====================================
\begin{lems} \label{lemma:fifofinalts}
If a process $p$ \cms{}s $m$, then $m'$, then the final values of their timestamps will be such that $m.ts < m'.ts$.
\end{lems}

\begin{proof}
From lines \ref{algline:gettime} and \ref{algline:rmlocal}, and the properties of the \rmc{} primitive, we know that all correct processes deliver $m$ and then $m'$, which means that each process $p'$ from $group(p)$ also inserts them into its $propPending$ list in this same order.

Suppose, by way of contradiction, that $m'$ is decided before $m$. This implies that some process $p$ proposed a message set that included $m'$, but not $m$. As $p$ inserted $m$ before $m'$ in its $propPending$ list, it means that $m$ was removed by $p$ from such list before being included in some proposed message set. However, the only possibility of removing a message from $propPending$ is on line \ref{algline:removedecided}, which means that $m$ already belonged to the $decided$ set, so it had been decided before $m'$, which is a contradiction. Therefore, $m'$ is not decided before $m$.

This also means that $m'$ is not inserted into the $decided$ list before $m$ has already been inserted. From line \ref{algline:gettime}, we know that the initial timestamp of $m$ is already smaller than that of $m'$. The only way to change this order is by making the final timestamp $m.ts$ bigger than $m'.ts$ on line \ref{algline:changetsafterdecision}. Even if this happens, as $m'$ is inserted into $decided$ after $m$, from line \ref{algline:conditionchangets}, we know that the final timestamp of $m'$ will be made greater than that of $m$.
\end{proof}

%===                 ==============================
%===   PROPOSITION   ==============================
%===                 ==============================
\begin{props} \label{props:fifoorder}
If $p$ \cms{}s $m$ and then $m'$, then no process $p'$ in both $m.dst$ and $m'.dst$ delivers $m'$ before delivering $m$.
\end{props}

\begin{proof}
Immediate from Lemma \ref{lemma:respecttimestamps} and Lemma \ref{lemma:fifofinalts}.
\end{proof}
